### 2021-02-25 rad-icfp-2021/background.tex
change outline
outline paper in intro
cleaned up refs.bib

\section{Rudiments of AD}

\subsection{Rudiments of differentiation and dual numbers.}
Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
$\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.

@@

The two main ways in which it is performed in practice is by operator overloading \cite{} or by source code transformation. We focus our work on the latter.

\subsection{Reverse-mode AD}

AD comes in two main flavors: forward mode and reverse mode. 
Forward mode computes a directional derivative of the Jacobian while reverse-mode computes a directional derivative of the transpose Jaobian. 
This has a significant impact in practice as this means reverse mode can compute the whole gradient of a function $f:\RR^n\to \RR$ in one pass, whereas forward mode needs $n$ passes. 
Assuming that both have about the same complexity, this can have a huge impact in practice. 
Such functions are expremely commonly seen as loss functions in the context of machine learning for instance.

Reverse-mode computes the derivatives backward, using the symmetry of the chain rule.
This means the compute flow is reverted after the first pass to compute the value of the function and its intermediate values. 
These values appear in the chain rule so should be stored or recomputed.
These things make reverse-mode harder to implement efficiently, and especially harder to analyse and optimise. 
Traditional implementions use references.
For instance, let 
 \begin{align*}
     \trm\defeq &\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
 \end{align*}
 The imperative version of the reverse differentiation of $\trm$ is
	\begin{align*}
		&w_1=\var_1 * \var_2 \\
		&w_2=w_1 * \var_1 \\
        &\var_1':=0~;~\var_2':=0 \\
        &w_1':=0 \\
		&w_2':=1 \\
		&w_1'+= \var_1 * w_2'; \var_1'+= w_1*w_2' \\
		&\var_1'+= \var_2 * w_1'; \var_2'+=\var_1 * w_1'\\
		&\text{return }(\var_1',\var_2')  
	\end{align*}
The tangent part for every variable from the context $\var_i'$ and every intermediate variable $w_i'$ is initialized at $0$, 
except for the return variable $w_2$ for which  $w_2':=1$.

As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
This representation is notoriously inefficient and all the implementions of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.

In the next section, we show on a very simple language how we can use standard optimisations to turn this inefficient representation into a very efficient one which is still purely functional.
The remaining of the paper will generalise this idea and explore some consequences of this new approach.

>>>>

\section{Reverse-mode in Automatic Differentiation}
\label{sec:background}

% \subsection{Rudiments of differentiation and dual numbers.}
% Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
% $\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.

% To find $\nabla f$ in a compositional way, two generalizations are reasonable:
% \begin{itemize}
% \item We need both $f$ and $\nabla f$ when calculating $\nabla (f;g)$
% of a composition $f;g$, using the chain rule, so we are really interested in the pair $(f,\nabla f):\RR\to \RR\times \RR$;
% \item In building $f$ we will need to consider functions of multiple arguments, such as $+:\RR^2\to \RR$, and these functions should propagate derivatives.
% \end{itemize}
% Thus we are more generally interested in transforming a function $g:\RR^n\to \RR$ into a function
% $h:(\RR\times \RR)^n\to \RR\times \RR$ in such a way that for any
% $f_1\dots f_n:\RR\to\RR$, 
% \begin{equation}
%   \label{eqn:dualnumber}
%   (f_1,\nabla f_1,\dots, f_n,\nabla f_n);h
%   =
%   ((f_1,\dots, f_n);g,\nabla ((f_1, \dots, f_n);g))\text.
% \end{equation}

% An intuition for $h$ is often given in terms of dual numbers.
% The transformed function operates on pairs of numbers, $(x,x')$, and it is common
% to think of such a pair as $x+x'\epsilon$ for an `infinitesimal' $\epsilon$.
% But while this is a helpful intuition, the formalization of infinitesimals can be intricate, 
% and the development in this paper is focussed on the elementary formulation in~\eqref{eqn:dualnumber}.

% The reader may also notice that $h$ encodes all the partial derivatives of
% $g$. For example, 
% if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
% $h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
% and similarly 
% $h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
% And conversely, if $g$ is differentiable in each argument, then
% a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
% combinations of partial derivatives:
% \[\textstyle h(x_1,x_1',x_2,x_2')=(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))\text.\]

% In summary, the idea of differentiation with dual numbers is 
% to transform a differentiable function
% $g:\RR^n\to \RR$ to a function $h:\RR^{2n}\to \RR^2$ which captures~$g$ and all its partial derivatives. We packaged this up in~\eqref{eqn:dualnumber} as a sort-of invariant which is useful for building derivatives of compound functions $\RR\to\RR$ in a compositional way.
% The idea of automatic differentiation is to perform this transformation at the source code level.

% The two main ways in which it is performed in practice is by operator overloading \cite{} or by source code transformation. We focus our work on the latter.

% \subsection{Reverse-mode AD}

% AD comes in two main flavors: forward mode and reverse mode. 
% Forward mode computes a directional derivative of the Jacobian while reverse-mode computes a directional derivative of the transpose Jaobian. 
% This has a significant impact in practice as this means reverse mode can compute the whole gradient of a function $f:\RR^n\to \RR$ in one pass, whereas forward mode needs $n$ passes. 
% Assuming that both have about the same complexity, this can have a huge impact in practice. 
% Such functions are expremely commonly seen as loss functions in the context of machine learning for instance.

% Reverse-mode computes the derivatives backward, using the symmetry of the chain rule.
% This means the compute flow is reverted after the first pass to compute the value of the function and its intermediate values. 
% These values appear in the chain rule so should be stored or recomputed.
% These things make reverse-mode harder to implement efficiently, and especially harder to analyze and optimize. 
% Traditional implementations use references.
% For instance, let 
%  \begin{align*}
%      \trm\defeq &\letin{w_1}{\var_1 * \var_2}{\\&\letin{w_2}{w_1 * \var_1}{\\&w_2}}
%  \end{align*}
%  The imperative version of the reverse differentiation of $\trm$ is
% 	\begin{align*}
% 		&w_1=\var_1 * \var_2 \\
% 		&w_2=w_1 * \var_1 \\
%         &\var_1':=0~;~\var_2':=0 \\
%         &w_1':=0 \\
% 		&w_2':=1 \\
% 		&w_1'+= \var_1 * w_2'; \var_1'+= w_1*w_2' \\
% 		&\var_1'+= \var_2 * w_1'; \var_2'+=\var_1 * w_1'\\
% 		&\text{return }(\var_1',\var_2')  
% 	\end{align*}
% The tangent part for every variable from the context $\var_i'$ and every intermediate variable $w_i'$ is initialized at $0$, 
% except for the return variable $w_2$ for which  $w_2':=1$.

% As noted in \cite{pearlmutter2008reverse}, in a functional language, one can have a purely functional reverse-mode using continuations. 
% This representation is notoriously inefficient, and the existing implementations of reverse mode in a functional setting have followed the lines of the imperative version above \cite{pearlmutter2008reverse, wang2018demystifying, baydin2016diffsharp}.

% In the next section, we first introduce a very simple first-order language. 
% Then we define reverse-mode on this language as a program transformation. 
% This first transformation is quite inefficient and we how we can use standard optimizations on functional languages to obtain a provably efficient pure program.
% The key idea behind that will be generalized and explored in detail in the rest of the paper is that of turning every operator into a unary one.     
% The remaining of the paper will generalize this idea and explore some consequences of this new approach.

\subsection{Rudiments of forward-mode AD and dual numbers}

Recall that the derivative of a function $f:\RR\to \RR$, if it exists, is a function
$\nabla f:\RR\to \RR$ such that $\nabla f(x_0)=\frac {\dif f(x)}{\dif x}(x_0)$ is the gradient of $f$ at $x_0$.

@@

The two main ways in which it is performed in practice is by operator overloading or by source code transformation (see e.g. \cite{griewank2008evaluating} Chapter 6). 
Our work focus on a compiled version so we focus on source code transformation which is more fitted in this case.

\subsection{Reverse-mode differentiation}

The problem with the previous approach shows up when one wants to compute the full gradient of a function $\RR^n\to\RR$, for a large $n$. 
Forward-mode only computes one directional derivative, for instance one partial derivative. This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
Suppose given a function $f=f_n\circ...\circ f_1 :\RR^n\to\RR$. 
The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(...(Jf_1v))...)$ for a direction $v\in\RR^n$. 
Reverse-mode, on the other hand, computes $(Jf)^Tv =Jf_1T^(Jf_{2}^T(...(Jf_n^Tv))...)$ for a vector $v\in\RR$. 
Because the compute flow of the function is reversed, the actual implementation of reverse-mode is way more tricky. 
%%TODO: from Matthijs' paper, need to rewrite.
Reverse-mode AD is only well-understood as a source-code transformation (also called define-then-run
style AD) on limited programming languages. Typically, its implementations
on more expressive languages that have features such as higher-order functions and conditionals
make use of define-by-run approaches. These approaches first build a computation graph during runtime, effectively evaluating the program until a straight-line
first-order program is left, and then they evaluate this new program \cite{carpenter2015stan,paszke2017automatic}. 
Such approaches have the severe downside that the differentiated code cannot benefit from existing optimizing compiler architectures. As such, these AD libraries
need to be implemented using carefully, manually optimized code, that for example does not contain any common subexpressions. This implementation process
is precarious and labour intensive. Further, some whole-program optimizations
that a compiler would detect go entirely unused in such systems.

\subsection{Purely functional inefficient reverse-mode}

Following \cite{pearlmutter2008reverse}, there is an easy way to define an inefficient yet purely functional reverse-mode transformation for first-order programs.
\MH{TODO: finish}

\subsection{Insights for efficient reverse-mode}

\begin{itemize}
    \item a
    \item  
\end{itemize}
